Types from data: \textnormal{Making structured data first-class citizens in F\#}

### 2015-11-20 pldi-2016/paper.tex
Final edits in section 4 (excluding examples)

\newcommand{\tsep}[0]{\; \triangledown \;}
\newcommand{\tytag}{\ident{tag}}
\newcommand{\dropopt}[1]{\lfloor#1\rfloor}
\newcommand{\addopt}[1]{\lceil#1\rceil}
\newcommand{\tytagof}{\ident{tagof}}
\newcommand{\nameoftag}{\ident{nameof}}

@@

\begin{figure}
\begin{center}
\includegraphics[scale=0.80,trim=5mm 5mm 5mm 5mm,clip]{images/hierarchy.pdf} % left bottom right top
\end{center}
\vspace{-0.5em}
\caption{Subtype relation between structural types}
\label{fig:subtyping-diagram}
\vspace{-0.5em}
\end{figure}

@@

In this section, we build the theoretical framework for proving relativised type safety of 
structural type providers. We discuss the runtime representation of documents and 
primitive operations (\S\ref{sec:formal-convert}), we embed these into a formal model of an F\# 
subset (\S\ref{sec:formal-ff}) and we describe how structural type providers turn
inferred structural types into F\# types (\S\ref{sec:formal-tp}).

@@

\begin{figure*}
\noindent
\begin{equation*}
\begin{array}{l}
\ident{asFloat}(i) \,\reduce f\quad(f = i)\\
\ident{asFloat}(d) \reduce f\quad(f = d)\\
\ident{asFloat}(f) \reduce f
\\[0.5em]
\ident{asDec}(i) \,\reduce d\quad(d = i)\\
\ident{asDec}(d) \reduce d
\end{array}
\qquad
\begin{array}{l}
\ident{hasType}(\nu \{ \nu_1 \!:\! \sigma_1, \ldots, \nu_n \!:\! \sigma_n \}, \nu'_? \{ \nu'_1\mapsto s_1, \ldots, \nu'_m\mapsto s_m \}) \reduce (\nu = \nu'_?) ~\wedge \\
  \quad (~ ((\nu_1 = \nu'_1) \wedge \ident{hasType}(\sigma_1, s_1)) \vee\ldots\vee ((\nu_1 = \nu'_m) \wedge \ident{hasType}(\sigma_1, s_m)) \vee \ldots \vee\\
  \quad ~\; ((\nu_n = \nu'_1) \wedge \ident{hasType}(\sigma_n, s_1)) \vee\ldots\vee ((\nu_n = \nu'_m) \wedge \ident{hasType}(\sigma_n, s_m))~)
\\[0.5em]
\ident{hasType}([\sigma], [s_1; \ldots; s_n]) \reduce \ident{hasType}(\sigma, s_1)\wedge\ldots\wedge\ident{hasType}(\sigma, s_n) \\
\ident{hasType}([\sigma], \kvd{null}) \reduce \kvd{true} \\  
\end{array}  
\end{equation*}
%
\vspace{-0.5em}
%
\begin{equation*}
\begin{array}{l}
\ident{isNull}(\kvd{null}) \reduce \kvd{true} \\
\ident{isNull}(\_) \reduce \kvd{false} 
\\[0.5em]
\ident{getField}(\nu,\nu_i, \nu~\{\ldots, \nu_i=s_i, \ldots\}) \reduce s_i\\
\ident{getField}(\bullet, \nu_i, \{\ldots, \nu_i=s_i, \ldots\}) \reduce s_i\\
\ident{getField}(\nu,\nu', \nu~\{\ldots, \nu_i=s_i, \ldots\}) \reduce \kvd{null}\quad(\nexists i.\nu_i=\nu' )\\
\ident{getField}(\bullet, \nu', \{\ldots, \nu_i=s_i, \ldots\}) \reduce \kvd{null}\quad(\nexists i.\nu_i=\nu' )
\\[0.5em]
\ident{getChildren}([s_1; \ldots; s_n]) \reduce [s_1; \ldots; s_n]  \\  
\ident{getChildren}(\kvd{null}) \reduce [~] 
\end{array}
~
\begin{array}{l}
\ident{hasType}(\ident{string}, t) \reduce \kvd{true} \\
\ident{hasType}(\ident{bool}, s) \reduce \kvd{true} \quad(\textnormal{when}~s\in{\kvd{true},\kvd{false}} )\\
\ident{hasType}(\ident{float}, s) \reduce \kvd{true} \quad(\textnormal{when}~s=i, s=d ~\textnormal{or}~ s=f) \\
\ident{hasType}(\ident{decimal}, s) \reduce \kvd{true} \quad(\textnormal{when}~s=i ~\textnormal{or}~ s=f) \\
\ident{hasType}(\ident{int}, i) \reduce \kvd{true}\\
\ident{hasType}(\kvd{option}\langl\sigma\rangl, s) \reduce \ident{hasType}(\sigma,s) \\
\ident{hasType}(\kvd{option}\langl\sigma\rangl, s) \reduce \kvd{true} \\
\ident{hasType}(\kvd{any}\langl\ldots\rangl, s) \reduce \kvd{true} \\
\ident{hasType}(\_, \_) \reduce \kvd{false} \\
\end{array}
\end{equation*}

@@

@@

\subsection{Structural values and conversions}
\label{sec:formal-convert}

As indicated by the subtyping, our system permits certain runtime conversions (e.g.~an integer
\num{1} can be treated as a floating-point \num{1.0}). The following primitive operations 
implement the conversions and other helpers:
%
\begin{equation*}
\begin{array}{lcl}
 op  &\narrow{=}& \ident{asDec}(s) \lsep \ident{asFloat}(s) \lsep \ident{getChildren}(s) \\
     &\narrow{|}& \ident{getField}(\nu?, \nu, s) \lsep \ident{isNull}(s) \lsep \ident{hasType}(\sigma, s)
\end{array}
\end{equation*}
%
The operations are used in code generated by type providers. Their behavior is defined
by the rules in Figure~\ref{fig:op-conversions}. The conversion functions (\ident{asDec}
and \ident{asFloat}) only perform conversions required by the subtyping (our actual 
implementation is more lax and performs additional conversions).

The \ident{getField} operation is used to access a field of a record. The operation ensures that
the actual record name matches the expected name (we write $\bullet$ for the name of an unnamed 
record). The \ident{getChildren} operation returns elements of a list and turns \kvd{null} into 
an empty collction.

Finally, we also define two helper functions. The \ident{isNull} operation is used to check whether 
value is \kvd{null} and \ident{hasType} is a runtime type test. The most complex case is handling 
of records where we check that for each field $\nu_1, \ldots, \nu_n$ in the type, the actual record 
value has a field of the same name with a matching type. The last line defines a ``catch all'' 
pattern, which returns \kvd{false} for all remaining cases. Although not elegant, the rules 
can be expressed using just Featherweight F\#, discussed in the next section.

@@

\begin{figure}
\noindent
\begin{equation*}
\begin{array}{rl}
 \textnormal{\footnotesize{(member)}}&
 \hspace{-0.4em}
 \inference
 { \kvd{type}~C(\overline{x:\tau})=\ldots \kvd{member}~N_i : \tau_i = e_i \ldots }
 { (\kvd{new}~C(\overline{v})).N_i \reduce e_i[\overline{x} \leftarrow \overline{v}] }\\
 \\
 \textnormal{\footnotesize{(eq1)}}&
 v=v\reduce\kvd{true} \qquad \textnormal{\footnotesize{(eq2)}}~~v=v'\reduce\kvd{false}\\
 \\
 \textnormal{\footnotesize{(cond1)}}&
 \hspace{-0.4em}
 \kvd{if}~\kvd{true}~\kvd{then}~e_1~\kvd{else}~e_2 ~\reduce~ e_1 \\
 \\
 \textnormal{\footnotesize{(cond2)}}&
 \hspace{-0.4em}
 \kvd{if}~\kvd{false}~\kvd{then}~e_1~\kvd{else}~e_2 ~\reduce~ e_2 \\
 \\
 \textnormal{\footnotesize{(match1)}}&
 \hspace{-1em}
 \begin{array}{l}
  \kvd{match}~\ident{None}~\kvd{with} \\
  \ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2
 \end{array} \hspace{-0.5em} ~\reduce~ e_2 \\
 \\
 \textnormal{\footnotesize{(match2)}}&
 \hspace{-1em}
 \begin{array}{l}
    \kvd{match}~\ident{Some}(v)~\kvd{with} \\
    \ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2
 \end{array} \hspace{-0.5em} ~\reduce~ e_1[x\leftarrow v]\\
 \\
 \textnormal{\footnotesize{(match3)}}&
 \hspace{-1em}
 \begin{array}{l}
  \kvd{match}~[v_1;\ldots;v_n]~\kvd{with} \\[0em]
  [x_1;\ldots;x_m ] \rightarrow e_1 \,|\, \_ \rightarrow e_2
 \end{array} \hspace{-0.5em} ~\reduce~ e_2\quad(m\neq n) \\
 \\
 \textnormal{\footnotesize{(match4)}}&
 \hspace{-1em}
 \begin{array}{l}
  \kvd{match}~[v_1;\ldots;v_n]~\kvd{with} \\[0em]
  [x_1;\ldots;x_n ] \rightarrow e_1 \,|\, \_ \rightarrow e_2
 \end{array} \hspace{-0.5em} ~\reduce~ e_1[\overline{x}\leftarrow\overline{v}] \\
 \\
 \textnormal{\footnotesize{(map)}}&
 \hspace{-0.4em}
 \ident{map}~(\lambda x.e)~[v_1; \ldots] ~\reduce~ [e[x\leftarrow v_1]; \ldots] \\
 \\
 \textnormal{\footnotesize{(ctx)}}&
 \hspace{-0.4em}
  E[e] \reduce E[e'] \qquad\qquad(\textnormal{when}~e \reduce e')\\
\end{array}
\end{equation*}

@@

\subsection{Featherweight F\#}
\label{sec:formal-ff}

We now extend the semantics discussed so far and add a minimal subset of F\# needed to model F\# 
Data. Type providers use classes and so we focus on the F\# object model and combine aspects of SML 
\cite{sml} and Featherweight Java \cite{fwjava}.

We only need classes with parameter-less members and without inheritance. A class has a single 
implicit constructor and the declaration closes over constructor parameters. To avoid including all 
of ML, we only pick constructs for working with options and lists that we need later.
%
\begin{equation*}
\begin{array}{rcl}
 \tau &\narrow{=}& \ident{int} \lsep \ident{decimal} \lsep \ident{float} \lsep \ident{bool} \lsep \ident{string} \\[0.0em]
      &\narrow{|}& C \lsep \ident{StructVal} \lsep \ident{list}\langl\tau\rangl \lsep \ident{option}\langl\tau\rangl \\[0.6em]
 L &\narrow{=}& \kvd{type}~C(\overline{x:\tau}) = \overline{M} \\[0.0em]
 M &\narrow{=}& \kvd{member}~N:\tau=e \\[0.6em]
 v &\narrow{=}& s \lsep \ident{None} \lsep \ident{Some}(v) \lsep \kvd{new}~C(\overline{v}) \lsep [v_1; \ldots; v_n] \\[0.0em]
 e &\narrow{=}& s \lsep op \lsep e.N \lsep \kvd{new}~C(\overline{e}) \lsep {\kvd{if}~e_1~\kvd{then}~e_2~\kvd{else}~e_3}\\
   &\narrow{|}& \ident{None} \lsep\kvd{match}~e~\kvd{with}~\ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2 \\
   &\narrow{|}& \ident{Some}(e) \lsep [e_1;\ldots;e_n]\lsep \ident{map}~(\lambda x\rightarrow e_1)~e_2\\
   &\narrow{|}& e = e \lsep \kvd{match}~e~\kvd{with}~[x_1; \ldots; x_n] \rightarrow e_1 \,|\, \_ \rightarrow e_2
\end{array}
\end{equation*}

\noindent
The type \ident{StructVal} is a type of all structural values $s$. A class
definition $L$ consists of a constructor and zero or more members. Values $v$ include 
previously defined structural values $s$ and values for the option and list type; finally 
expressions $e$ include previously defined operations $op$, class construction, member access, 
conditionals and expressions for working with option values and lists. We include 
\ident{map} as a special construct to avoid making the language too complex.

\paragraph{Reduction.}
The reduction relation is of the form $e \reduce e'$. We also write 
$e \reduce^{*} e'$ to denote the reflexive and transitive closure of $\reduce$. The reduction rules
for operations $op$ were discussed earlier. Figure~\ref{fig:ff-reduction} shows the remaining
rules.

The (\emph{ctx}) rule performs a reduction inside a sub-expression specified by an evaluation context.
This models the eager evaluation of F\#. An evaluation context $E$ is defined as:
%
\begin{equation*}
\begin{array}{rcl}
 E &\narrow{=}& [\overline{v};E;\overline{e}] \lsep E.N \lsep \kvd{new}~C(\overline{v}, E, \overline{e})\\[0.1em]
   &\narrow{|}& \ident{map}~(\lambda x\rightarrow e)~E \lsep \kvd{if}~E~\kvd{then}~e_1~\kvd{else}~e_2 \\[0.1em]
   &\narrow{|}& \ident{Some}(E) \lsep op(E) \lsep E = e \lsep v = E \\[0.1em]
   &\narrow{|}& \kvd{match}~E~\kvd{with}~\ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2 \\[0.1em]
   &\narrow{|}& \kvd{match}~E~\kvd{with}~[x_1; \ldots; x_n] \rightarrow e_1 \,|\, \_ \rightarrow e_2
\end{array} 
\end{equation*}

\noindent
The evaluation first reduces arguments of functions and the evaluation proceeds from left to right 
as denoted by $\overline{v}, E, \overline{e}$ in constructor arguments or $\overline{v};E;\overline{e}$
in list initialization.

We write $e[\overline{x} \leftarrow \overline{v}]$ for the result of replacing variables $\overline{x}$ by
values $\overline{v}$ in an expression. The (\emph{member}) rule reduces a member access using a class 
definition in the assumption to obtain the body of a member. The remaining six rules
give standard reductions for conditionals and pattern matching.

We omit expressions $e \vee e$ and $e \wedge e$ that are used in Figure~\ref{fig:op-conversions}, 
as those can be easily defined as syntactic sugar using \kvd{if}. All expressions reduce to a value in a 
finite number of steps or get stuck due to an error condition. An error condition can be a wrong 
argument passed to conditional, pattern matching or one of the primitives from Figure~\ref{fig:op-conversions}.

@@

\begin{figure}
\noindent  
\begin{equation*}
\inference
  {L; \Gamma \vdash e_1 : \ident{StructVal}}
  {L; \Gamma \vdash [e_1; \ldots; e_n]:\ident{StructVal}}
\quad  
\inference
  {~}
  {L; \Gamma \vdash n : \ident{int}}
\end{equation*}
\begin{equation*}
\inference
  {L; \Gamma \vdash e_1 : \tau}
  {L; \Gamma \vdash [e_1; \ldots; e_n]:\ident{list}\langl\tau\rangl}
\quad
\inference
  {~}
  {L; \Gamma \vdash s : \ident{StructVal}}
\end{equation*}
\vspace{0.5em}
\begin{equation*}
\inference
  {L; \Gamma \vdash e : C \\ \kvd{type}~C(\overline{x:\tau}) = ..\;\kvd{member}~N_i : \tau_i = e_i\;.. \in L}
  {L; \Gamma \vdash e.N_i:\tau_i}
\end{equation*}
\vspace{0.25em}
\begin{equation*}
\inference
  {L; \Gamma \vdash e_i : \tau_i & \kvd{type}~C(x_1:\tau_1, \ldots, x_n:\tau_n) = \ldots \in L}
  {L; \Gamma \vdash \kvd{new}~C(e_1, \ldots, e_n):C}
\end{equation*}

@@

So far, we defined the type inference algorithm which produces a structural type $\sigma$ from one 
or more sample documents (\S\ref{sec:inference}) and we defined a simplified model of evaluation
of F\# (\S\ref{sec:formal-ff}) and F\# Data runtime (\S\ref{sec:formal-tp}). In this section, we 
define how the type providers work, linking the two parts.

All F\# Data type providers take (one or more) sample documents, infer a common supertype $\sigma$
and then use it to generate F\# types that are exposed to the programmer.\footnote{The actual 
implementation provides \emph{erased types} as described in \cite{fsharp-typeprov}. Here, we treat 
the code as actually generated. This is an acceptable simplification, because F\# Data type providers 
do not rely on laziness that is available through erased types.}

>>>>

\newcommand{\tsep}[0]{\; \triangledown \;}
\newcommand{\tytag}{\ident{tag}}
\newcommand{\dropopt}[1]{\lfloor#1\rfloor}
\newcommand{\addopt}[1]{\lceil#1\rceil}
\newcommand{\tytagof}{\ident{tagof}}

@@

\begin{figure}
\begin{center}
\includegraphics[scale=0.80,trim=5mm 5mm 5mm 5mm,clip]{images/hierarchy.pdf} % left bottom right top
\end{center}
\vspace{-0.5em}
\caption{Important aspects of the preferred shape relation}
\label{fig:subtyping-diagram}
\vspace{-0.5em}
\end{figure}

@@

This section presents the theoretical model of strucutral type providers. We introduce the Foo calculus 
which is an object-oriented subset of F\# based on SML \cite{sml} and Featherweight Java \cite{fwjava}. It 
includes operations for working with data that model the F\# Data runtime. Finally, we describe how 
type providers turn inferred shapes into Foo classes (\S\ref{sec:formal-tp}).

Type providers map the dirty world of data into a nice world of F\#. To model this, the Foo calculus
does not have \kvd{null} values and data values $d$ are never directly exposed.

@@

\begin{figure*}
\noindent
\begin{equation*}
\hspace{1.1em}
\begin{array}{l}
\ident{hasShape}(\nu~ \{ \nu_1 \!:\! \sigma_1, \ldots, \nu_n \!:\! \sigma_n \}, \nu'~\{ \nu'_1\mapsto d_1, \ldots, \nu'_m\mapsto d_m \}) \reduce (\nu = \nu') ~\wedge \\
  \quad (~ ((\nu_1 = \nu'_1) \wedge \ident{hasShape}(\sigma_1, d_1)) \vee\ldots\vee ((\nu_1 = \nu'_m) \wedge \ident{hasShape}(\sigma_1, d_m)) \vee \ldots \vee\\
  \quad ~\; ((\nu_n = \nu'_1) \wedge \ident{hasShape}(\sigma_n, d_1)) \vee\ldots\vee ((\nu_n = \nu'_m) \wedge \ident{hasShape}(\sigma_n, d_m))~)
\\[0.5em]
\ident{hasShape}([\sigma], [d_1; \ldots; d_n]) \reduce \ident{hasShape}(\sigma, d_1)\wedge\ldots\wedge\ident{hasShape}(\sigma, d_n) \\
\ident{hasShape}([\sigma], \kvd{null}) \reduce \kvd{true} 
\end{array}  
\quad
\begin{array}{l}
\ident{convFloat}(\ident{float}, i) \reduce f~(f=i) \\
\ident{convFloat}(\ident{float}, f) \reduce f \\[0.75em]
\ident{convNull}(\kvd{null}, e) \reduce \ident{None} \\
\ident{convNull}(d, e) \reduce \ident{Some}(e~d) 
\end{array}
\end{equation*}
%
\vspace{-0.75em}
%
\begin{equation*}
\begin{array}{l}
\ident{hasShape}(\ident{string}, s) \reduce \kvd{true} \\
\ident{hasShape}(\ident{int}, i) \reduce \kvd{true}\\
\ident{hasShape}(\ident{bool}, d) \reduce \kvd{true} \quad(\textnormal{when}~d\in{\kvd{true},\kvd{false}} )\\
\ident{hasShape}(\ident{float}, d) \reduce \kvd{true} \quad(\textnormal{when}~d=i ~\textnormal{or}~ d=f) \\
\ident{hasShape}(\_, \_) \reduce \kvd{false} \\
\end{array}
\;\;
\begin{array}{l}
\ident{convPrim}(\sigma, d) \reduce d\quad(\sigma,d\in\{ (\ident{int},i),(\ident{string},s),(\ident{bool}, b) \})
\\[0.25em]
\ident{convField}(\nu,\nu_i, \nu~\{\ldots, \nu_i=d_i, \ldots\}, e) \reduce e~d_i\\
\ident{convField}(\nu,\nu', \nu~\{\ldots, \nu_i=d_i, \ldots\}, e) \reduce e~\kvd{null}\quad(\nexists i.\nu_i=\nu' )
\\[0.25em]
\ident{convElements}([d_1; \ldots; d_n], e) \reduce e~d_1 :: \ldots :: e~d_n :: \kvd{nil}  \\  
\ident{convElements}(\kvd{null}) \reduce \kvd{nil}
\end{array}
\end{equation*}

@@

\subsection{The Foo calculus}
\label{sec:formal-ff}

The syntax of the calculus is shown in Figure~\ref{fig:foo-syntax}. We only need classes with 
parameter-less members and without inheritance. A class has a single implicit constructor and the 
declaration closes over constructor parameters.

The type \ident{Data} is the type of all structural data $d$. A class definition $L$ consists of 
a constructor and zero or more members. Values $v$ include previously defined data $d$; expressions
$e$ include class construction, member access and the usual functional constructs (functions,
lists, options) and conditionals. The $op$ constructs are discussed next.

@@

\paragraph{Dynamic data operations.}

The Foo programs can only work with raw data using certain primitive operations. Those are modelled
by the $op$ primitives. In F\# Data, those are internal and users never access them directly.

@@

\begin{figure}[!h]
\noindent
\begin{equation*}
\begin{array}{rcl}
 \tau &\narrow{=}& \ident{int} \lsep \ident{float} \lsep \ident{bool} \lsep \ident{string} \lsep C \lsep \ident{Data} \\[0.0em]
      &\narrow{|}& \tau_1 \rightarrow \tau_2 \lsep \ident{list}\langl\tau\rangl \lsep \ident{option}\langl\tau\rangl 
\\[0.6em]
 L &\narrow{=}& \kvd{type}~C(\overline{x:\tau}) = \overline{M} \\[0.0em]
 M &\narrow{=}& \kvd{member}~N:\tau=
\\[0.6em]
 v &\narrow{=}& d \lsep \ident{None} \lsep \ident{Some}(v) \lsep \kvd{new}~C(\overline{v}) \lsep v_1::v_2 \\[0.0em]
 e &\narrow{=}& d \lsep op \lsep e_1~e_2 \lsep \lambda x.e \lsep e.N \lsep \kvd{new}~C(\overline{e}) \\
   &\narrow{|}& \ident{None} \lsep\kvd{match}~e~\kvd{with}~\ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2 \\
   &\narrow{|}& \ident{Some}(e) \lsep e_1 = e_2 \lsep {\kvd{if}~e_1~\kvd{then}~e_2~\kvd{else}~e_3} \lsep \kvd{nil} \\
   &\narrow{|}& e_1 :: e_2 \lsep \kvd{match}~e~\kvd{with}~x_1::x_2 \rightarrow e_1 \,|\, \kvd{nil} \rightarrow e_2 
\\[0.6em]
op &\narrow{=}& \ident{convFloat}(\sigma, e) \lsep \ident{convPrim}(\sigma, e) \\
   &\narrow{|}& \ident{convField}(\nu_1, \nu_2, e, e) \lsep \ident{convNull}(e_1, e_2) \\
   &\narrow{|}& \ident{convElements}(e_1, e_2) \,\lsep \ident{hasShape}(\sigma, e) 
\end{array}
\end{equation*}

\caption{The syntax of the Foo calculus}
\label{fig:foo-syntax}
\vspace{-0.5em}
\end{figure}

The behaviour of the dynamic data operations is defined by the reduction rules in 
Figure~\ref{fig:op-conversions}. The operations can convert data values into values of 
less preferred shape. For example, \ident{convFloat} turns an integer \num{1} into 
a floating-point \num{1.0} and \ident{convElements} turns \kvd{null} value into an empty list.

All of the dynamic data operations take a data value $d$ and produce a Foo value. We do not 
make the conversion explicit for primitive values (\ident{convPrim}) and assume that primitive
values are shared, but we convert all other values. For example \ident{convElements} turns a 
data collection $[d_1, \ldots, d_n]$ into a Foo list $v_1 :: .. :: v_n :: \kvd{nil}$.

The data operations for collections, record fields and nullable values all take a function
as the last argument and invoke it to convert the obtained data. For example, \ident{convField} 
accesses a record field -- it ensures that the actual record name matches the expected name and 
passes the field value (or \kvd{null} data value) to the provided function.

Finally, we also define a runtime shape test \ident{hasShape}.
The most complex case is handling of records where we check that for each field $\nu_1, \ldots, \nu_n$ 
in the record, the actual record value has a field of the same name with a matching shape. The last 
line defines a ``catch all'' pattern, which returns \kvd{false} for all remaining cases. Although
not elegant, if we treat $e_1 \vee e_2$ and $e_1 \wedge e_2$ as syntactic sugar for \kvd{if} then
the rule is expressed using just Foo expressions.

\paragraph{Reduction.}
The reduction relation is of the form $L, e \reduce e'$. We often omit class declarations 
$L$ and we write $L, e \reduce^{*} e'$ to denote 
the reflexive and transitive closure of $\reduce$.

Figure~\ref{fig:ff-reduction} shows the reduction rules. The (\emph{ctx}) rule models the eager
evaluation of F\# and performs a reduction inside a sub-expression specified by an evaluation 
context $E$:
%
\begin{equation*}
\begin{array}{rcl}
 E &\narrow{=}& v::E \lsep v~E \lsep E.N \lsep \kvd{new}~C(\overline{v}, E, \overline{e})\\[0.1em]
   &\narrow{|}&  \kvd{if}~E~\kvd{then}~e_1~\kvd{else}~e_2  \lsep E = e \lsep v = E \\[0.1em]
   &\narrow{|}& \ident{Some}(E) \lsep op(\overline{v}, E, \overline{e})\\[0.1em]
   &\narrow{|}& \kvd{match}~E~\kvd{with}~\ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2 \\[0.1em]
   &\narrow{|}& \kvd{match}~E~\kvd{with}~x_1 :: x_2 \rightarrow e_1 \,|\, \kvd{nil} \rightarrow e_2
\end{array} 
\end{equation*}

\noindent
The evaluation proceeds from left to right as denoted by $\overline{v}, E, \overline{e}$ in 
constructor and dynamic data operation arguments or $v::E$ in list initialization.

We write $e[\overline{x} \leftarrow \overline{v}]$ for the result of replacing variables $\overline{x}$ by
values $\overline{v}$ in an expression. The (\emph{member}) rule reduces a member access using a class 
definition in the assumption to obtain the body of a member. The remaining six rules
give standard reductions for a functional language.

\paragraph{Type checking.} 
Well-typed Foo programs reduce to a value in a finite number of steps or get stuck due to an
error condition. The stuck states can only be due to the dynamic data operations (e.g.
$\ident{convFloat}(\ident{float},\kvd{null})$). The relativized type safety (Theorem~\ref{thm:safety})
characterizes the additional conditions on input data under which Foo programs do not ``go wrong''.

Typing rules in Figure~\ref{fig:ff-typecheck} are written using a judgement
$L; \Gamma \vdash e : \tau$ where the context also contains a set of class declarations $L$.
The fragment demonstrates the differences and similarities with SML and Featherweight Java:
%
\begin{itemize}
\item[--] All data values $d$ have the type \ident{Data}, but primitive data values (Booleans, 
  strings, integers and floats) can be implicitly converted to Foo values and so they also have a 
  primitive type as illustrated by the rule for $i$ and $f$.
\item[--] For non-primitive data values (including \kvd{null}, data collections and records),
  \ident{Data} is the only type.
\item[--] Operations $op$ (omitted) have a family of types (Doo does not have polymorphic types),
  accepting \ident{Data} as one of the arguments and producing a non-data Foo type.
\item[--] Rules for checking class construction and member access are similar to corresponding
  rules of Featherweight Java.  
\end{itemize}
%
An important part of Featherweight Java that is omitted here is the checking of type declarations
(ensuring the members are well-typed). We consider only classes generated by our type providers 
and those are well-typed by construction.

\begin{figure}
\noindent
\begin{equation*}
\begin{array}{rl}
 \textnormal{\footnotesize{(member)}}&
 \hspace{-0.6em}
 \inference
 { \kvd{type}~C(\overline{x:\tau})= \kvd{member}~N_i : \tau_i = e_i \ldots \in L }
 { L, (\kvd{new}~C(\overline{v})).N_i \reduce e_i[\overline{x} \leftarrow \overline{v}] }\\
 \\[-0.2em]
 \textnormal{\footnotesize{(match1)}}&
 \hspace{-1em}
 \begin{array}{l}
  \kvd{match}~\ident{None}~\kvd{with} \\
  \ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2
 \end{array} \hspace{-0.5em} ~\reduce~ e_2 \\
 \\[-0.2em]
 \textnormal{\footnotesize{(match2)}}&
 \hspace{-1em}
 \begin{array}{l}
    \kvd{match}~\ident{Some}(v)~\kvd{with} \\
    \ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2
 \end{array} \hspace{-0.5em} ~\reduce~ e_1[x\leftarrow v]\\
 \\[-0.2em]
 \textnormal{\footnotesize{(match3)}}&
 \hspace{-1em}
 \begin{array}{l}
  \kvd{match}~\kvd{nil}~\kvd{with} \\[0em]
  x_1::x_2 \rightarrow e_1 \,|\, \kvd{nil} \rightarrow e_2
 \end{array} \hspace{-0.5em} ~\reduce~ e_2\\
 \\[-0.2em]
 \textnormal{\footnotesize{(match4)}}&
 \hspace{-1em}
 \begin{array}{l}
  \kvd{match}~v_1::v_2~\kvd{with} \\[0em]
  x_1::x_2 \rightarrow e_1 \,|\, \kvd{nil} \rightarrow e_2
 \end{array} \hspace{-0.5em} ~\reduce~ e_1[\overline{x}\leftarrow\overline{v}] \\
 \\[-0.2em]
 \textnormal{\footnotesize{(cond1)}}&
 \hspace{-0.4em}
 \kvd{if}~\kvd{true}~\kvd{then}~e_1~\kvd{else}~e_2 ~\reduce~ e_1 \\
 \\[-0.2em]
 \textnormal{\footnotesize{(cond2)}}&
 \hspace{-0.4em}
 \kvd{if}~\kvd{false}~\kvd{then}~e_1~\kvd{else}~e_2 ~\reduce~ e_2 \\
 \\[-0.3em]
 \textnormal{\footnotesize{(eq1)}}&
 v=v\reduce\kvd{true} \qquad \textnormal{\footnotesize{(eq2)}}~~v=v'\reduce\kvd{false}\\
 \\[-0.3em]
 \textnormal{\footnotesize{(fun)}}&
 \hspace{-0.4em}
 (\lambda x.e)~v ~\reduce~ e[x\leftarrow v] \\
 \\[-0.3em]
 \textnormal{\footnotesize{(ctx)}}&
 \hspace{-0.4em}
  E[e] \reduce E[e'] \qquad\qquad(\textnormal{when}~e \reduce e')\\
\end{array}
\end{equation*}

@@

So far, we defined the type inference algorithm which produces a shape $\sigma$ from one 
or more sample documents (\S\ref{sec:inference}) and we defined a simplified model of evaluation
of F\# (\S\ref{sec:formal-ff}) and F\# Data runtime (\S\ref{sec:formal-tp}). In this section, we 
define how the type providers work, linking the two parts.

All F\# Data type providers take (one or more) sample documents, infer a common preferred shape $\sigma$
and then use it to generate F\# types that are exposed to the programmer.\footnote{The actual 
implementation provides \emph{erased types} as described in \cite{fsharp-typeprov}. Here, we treat 
the code as actually generated. This is an acceptable simplification, because F\# Data type providers 
do not rely on laziness that is available through erased types.}

\paragraph{Type provider mapping.}
A type provider produces an F\# type $\tau$ together with a Foo expression and a collection of 
class definitions. We express it using the following mapping:
%
\begin{equation*}
\sem{-} : \sigma \rightarrow (\tau \times e \times L) \qquad (\textnormal{where}~L,\emptyset \vdash e:\ident{Data}\rightarrow\tau)
\end{equation*}
%
The mapping $\sem{\sigma}$ takes an inferred shape $\sigma$. It returns an F\# type $\tau$ and
a function that turns the input data (value of type \ident{Data}) into a Foo value of type $\tau$. 
The type provider also generates class definitions that may be used by $e$.

Figure~\ref{fig:tp-generation} defines $\sem{-}$. Primitive types are handled by a single rule that
inserts an appropriate conversion function; \ident{convPrim} just checks that the shape matches
and \ident{convFloat} converts numbers to a floating-point.

For records, we generate a class $C$ that takes a data value as constructor parameter. For each 
record field, we generate a member with the same name as the field.\footnote{The actual
F\# Data implementation also capitalizes the names.} The body of the member calls \ident{convField}
with a function obtained from $\sem{\sigma_i}$. This function turns the field value (data of shape
$\sigma_i$) into a Foo value of type $\tau_i$. The returned expression creates a new instance of 
$C$ and the mapping returns the class $C$ together with all recursively generated classes. Note that 
the class name $C$ is not directly accessed by the user and so we can use arbitrary name, although the 
actual implementation in F\# Data attempts to infer a reasonable name.\footnote{For example, in 
\ident{\{\str{person}:\{\str{name}:\str{Tomas}\}\}}, the nested record will be named \ident{Person}
based on the name of the parent record field.}

A collection shape becomes an Foo $\ident{list}\langl\tau\rangl$. The returned expression calls \ident{convElements}
(which returns empty list for data value \kvd{null}). The last parameter is the recursively obtained
conversion function for the shape of elements $\sigma$. The handling of the nullable shape is similar,
but uses \ident{convNull}.

As discussed earlier, labelled top shape are also generated as Foo classes with properties. Given 
$\kvd{any}\langl\sigma_1, \ldots, \sigma_n\rangl$, we get corresponding F\# types $\tau_i$ and generate 
$n$ members of type $\ident{option}\langl \tau_i\rangl$. When the member is accessed, we need to perform
a runtime shape test using \ident{hasShape} to ensure that the value has the right shape (similarly to runtime 
type conversions from the top type in languages like Java). If the shape matches, a \ident{Some} value is 
returned. The shape inference algorithm also guarantees that there is only one case for each shape tag 
(\S\ref{sec:inference-commonsuper}) and so we can use the tag for the name of the generated member.

@@

\begin{figure}
\noindent  
\begin{equation*}
\inference
  {~}
  {L; \Gamma \vdash d : \ident{Data}}
~~
\inference
  {~}
  {L; \Gamma \vdash i : \ident{int}}
~~
\inference
  {~}
  {L; \Gamma \vdash f : \ident{float}}
\end{equation*}
\vspace{0.25em}
\begin{equation*}
\inference
  {L; \Gamma, x : \tau_1 \vdash e : \tau_2}
  {L; \Gamma \vdash \lambda x.e : \tau_2}
\quad
\inference
  {L; \Gamma \vdash e_2 : \tau_1 \\
   L; \Gamma \vdash e_1 : \tau_1 \rightarrow \tau_2}
  {L; \Gamma \vdash e_1~e_2 : \tau_2}
\end{equation*}
\vspace{0.5em}
\begin{equation*}
\inference
  {L; \Gamma \vdash e : C \\ \kvd{type}~C(\overline{x:\tau}) = ..\;\kvd{member}~N_i : \tau_i = e_i\;.. \in L}
  {L; \Gamma \vdash e.N_i:\tau_i}
\end{equation*}
\vspace{0.25em}
\begin{equation*}
\inference
  {L; \Gamma \vdash e_i : \tau_i & \kvd{type}~C(x_1:\tau_1, \ldots, x_n:\tau_n) = \ldots \in L}
  {L; \Gamma \vdash \kvd{new}~C(e_1, \ldots, e_n):C}
\end{equation*}

@@
